---
title: "For_Amany"
author: "Tyler Muffly, MD"
date: "2023-12-28"
output: html_document
  pdf_document: default
  html_document:
    self_contained: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    message = FALSE,
    warning = FALSE,
    cache = FALSE,
    tidy = FALSE,
    fig.width = 7,
    fig.height = 7,
    include = FALSE,
    out.width = "65%")
```

This tutorial demonstrates how to do a drive time analysis in R. Drive time analyses can answer what areas lie within a specified driving time from a certain point, and what are the demographics of the people inside and outside that area.

The tutorial is adapted from [code](https://github.com/khnews/2021-delta-appalachia-stroke-access) used in the analysis for [In Appalachia and the Mississippi Delta, Millions Face Long Drives to Stroke Care](https://khn.org/news/article/appalachia-mississippi-delta-stroke-treatment-advanced-care-rural-access/). The story was a collaboration between KHN and InvestigateTV and published in May, 2021.

The basic process we'll follow is:

1.  Get a dataset of locations of interest

2.  Retrieve a drive-time isochrone for each point

3.  Calculate how much each Census block group in our area of interest overlaps with the isochrones

4.  Get population data for each block group

5.  Calculate how many people live within the isochrones and how many people live outside, using our block group data

### Requirements
This analysis requires  a [HERE API key](https://developer.here.com/) and a [U.S. Census Bureau API key](https://api.census.gov/data/key_signup.html) for data retrieval.

### Library Packages
We'll get GYN ONC locations.  
```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(tidyr)
library(sf)
library(leaflet)
library(tigris)
library(ggplot2)
library(hereR)
library(censusapi)
library(htmlwidgets)
library(webshot)
library(memoise)
library(tidyverse)
library(viridis)
library(wesanderson) # color palettes
library(mapview)
library(htmlwidgets)
library(shiny) # creation of GUI, needed to change leaflet layers with dropdowns
library(htmltools) #added for saving html widget
devtools::install_github('ramnathv/htmlwidgets')
library(htmlwidgets)
library (leaflet.extras)
library (leaflet.minicharts)
library(formattable)
library(tidycensus)

# Set tigris options for using cache
options(tigris_use_cache = TRUE)

# Initialize HERE API
hereR::set_freemium(ans = FALSE)
hereR::set_key(Sys.getenv("HERE_API_KEY"))
hereR::set_verbose(TRUE)
if (!nzchar(Sys.getenv("HERE_API_KEY"))) {
  stop("HERE_API_KEY environment variable is not set.")
}


# Add key to .Renviron. # Reload .Renviron
Sys.setenv(CENSUS_KEY = Sys.getenv("CENSUS_API_KEY"))
readRenviron("~/.Renviron")
```

# 1.  Get a dataset of locations of gynecologic oncologists from `GOBA`
### Retrieve isochrones with the `hereR` package 
Now that we have our set of GYN oncologists, we'll get the 60-minute drive time radius for each one using the [hereR](https://munterfinger.github.io/hereR/) package. This package wraps the [HERE Technologies](https://developer.here.com/) API. You'll need a developer key to use it. 

We'll use hereR's `isoline()` function to get the isochrones. This code defines a function to calculate isochrones for a given location and time, with a focus on healthcare access during flu season. It uses the `hereR` package to perform these calculations and handles potential errors or warnings during the process.

Once in a while you might have a server issue or some other error. I like to wrap the API function in a `tryCatch()` function so that if any one point fails the rest still go forward. Then, if any points did fail, I loop through the list that failed to try them again. If you're using a larger dataset I highly recommend having a setup like this so that you're not making a bunch of API hits only to hit an error and abort your whole script.

```{r}
gyn_onc <- readr::read_rds("data/references/_Recent_Grads_GOBA_NPI_2022.rds") %>%
  dplyr::filter(sub1=="ONC") %>%
  dplyr::rename(name = name.x) %>%
  dplyr::mutate(unique_id = dplyr::row_number())

gyn_onc_sf <- gyn_onc %>%
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326) %>%
  head(5)

# Initialize HERE API
hereR::set_freemium(ans = FALSE)
hereR::set_verbose(TRUE)
if (!nzchar(Sys.getenv("HERE_API_KEY"))) {
  stop("HERE_API_KEY environment variable is not set.")
}

```

```{r}
create_isochrones <- memoise::memoise(function(location, range, posix_time = as.POSIXct("2023-10-20 08:00:00", format = "%Y-%m-%d %H:%M:%S")) {


  if (!nzchar(Sys.getenv("HERE_API_KEY"))) {
    stop("HERE_API_KEY environment variable is not set.")
  }

  hereR::set_key(Sys.getenv("HERE_API_KEY"))


  cat("\033[Display setup instructions:\033[0m\n")
  cat("\033[34mTo create isochrones for a specific point(s) use the following code:\033[0m\n")
  cat("\033[34mtryLocationMemo(location = location, range = c(1800, 3600, 7200, 10800))\n")

  # Check if HERE_API_KEY is set in Renviron
  if (Sys.getenv("HERE_API_KEY") == "") {
    cat("Please set your HERE API key in your Renviron file using the following steps:\n")
    cat("1. Add key to .Renviron\n")
    cat("Sys.setenv(HERE_API_KEY = \"your_api_key_here\")\n")
    cat("2. Reload .Renviron\n")
    cat("readRenviron(\"~/.Renviron\")\n")
    stop("HERE_API_KEY environment variable is not set. Please set it to your HERE API key.")
  }

  # Initialize HERE API securely using an environment variable for the API key
  cat("Setting up the hereR access...\n")
  hereR::set_freemium(ans = FALSE)
  hereR::set_key(Sys.getenv("HERE_API_KEY"))

  hereR::set_verbose(TRUE)

  # Initialize a list to store the isolines
  isolines_list <- list()

  # Try to calculate isolines for the given location
  out <- tryCatch({
    for (r in range) {
      # Calculate isolines using hereR::isoline function
      temp <- hereR::isoline(
        poi = location, #sf object
        datetime = posix_time, #POSIXct object, datetime for the departure
        routing_mode = "fast", #Try to route fastest route or "short"est route.
        range = r,  # Time range in seconds
        range_type = "time", # character of the isolines: "distance" or "time"
        transport_mode = "car", #specified for "car" transport instead of "truck" or "pedestrian"
        url_only = FALSE,
        optimize = "balanced",
        traffic = TRUE, # Includes real-time traffic
        aggregate = FALSE
      )

      # Log the successful calculation
      cat("Isoline successfully produced for range:", r, "seconds\n")

      # Store the isoline in the list
      isolines_list[[as.character(r)]] <- temp
    }

    # Return the list of isolines
    return(isolines_list)
  }, error = function(e) {
    # Handle any errors that occur during the calculation
    cat("Error in tryLocationMemo:", e$message, "\n")

    # Return an error message as a list
    return(list(error = e$message))
  })

  # Return the result, whether it's isolines or an error message
  return(out)
  cat("\tryLocation complete.\n")
})
```


```{r}
test_create_isochrones_for_dataframe <- function(dataframe, breaks = c(0, 30, 60, 120, 180)) {
  
  # Initialize isochrones as an empty NULL
  isochrones <- list()
  
  # Loop over the rows in the dataframe
  for (i in 1:nrow(dataframe)) {
    print(i)

    # Get the point for the current row
    point_temp <- dataframe[i, ]
    
    # Get isochrones for that point
    Sys.sleep(0.4)
    isochrone_result <- create_isochrones(location = point_temp, range = c(1800, 3600, 7200, 10800))
    
    if (is.list(isochrone_result) && "error" %in% names(isochrone_result)) {
      # Handle the case where an error occurred
      cat("Error: ", isochrone_result$error, "\n")
      isochrones[[i]] <- "Error"
    } else {
      # Handle the case where isochrones were successfully generated
      isochrones[[i]] <- isochrone_result
    }
  }
  
  # Convert the list of isochrones to a data frame
  isochrones_df <- do.call(rbind, lapply(isochrones, function(x) {
    if (is.character(x) && x == "Error") {
      # Create a data frame with "Error" if no isochrone was generated
      data.frame(range = breaks, geometry = NA)
    } else {
      # Extract the relevant information from the isochrone result
      data.frame(range = x$range, geometry = x$geometry)
    }
  }))
  
  # Save the isochrones data to an RDS file
  readr::write_rds(isochrones_df, paste("isochrones_raw_output_from_here_api_", format(Sys.time(), format = "%Y-%m-%d_%H-%M-%S"), ".rds", sep = ""))
  
  print(isochrones_df)
}

# Create a data frame with three rows of coordinates at the North Pole
North_Pole_Coordinates <- data.frame(
  lat = c(90, 90, 90),
  lon = c(0, 120, -120)
)

North_pole <- test_create_isochrones_for_dataframe(North_Pole_Coordinates, breaks = c(0, 30, 60, 120, 180))

isochrones_df
class(isochrones_df)
dim(isochrones_df)
```


```{r}
isoChrone.build <- function(locationsCoords, res = 100, breaks) {
    # Check if locationsCoords is an sf object
  if (!inherits(locationsCoords, "sf")) {
    stop("locationsCoords must be of class 'sf'.")
  }
  
  # Check if locationsCoords has the required columns
  required_columns <- c("lat", "long")
  if (!all(required_columns %in% names(locationsCoords))) {
    stop("locationsCoords must have 'lat' and 'long' columns.")
  }
  
  # Initialize variables
  n_breaks <- length(breaks)
  num_locations <- nrow(locationsCoords)
  isoChrones <- vector("list", length = num_locations)
  isoChrones.errors <- vector("list", length = num_locations)
  
  # Main loop: Continue until all isochrones are generated or error limit is reached
  while (any(sapply(isoChrones, function(x) is.null(x)))) {
    cat("\f\n[ Building isoChrones ]\n\n")
    
    # Loop through each set of coordinates
    for (idx in 1:num_locations) {
      if (is.null(isoChrones[[idx]])) {
        cat("\rFacility (", idx, "/", num_locations, ") [Facility Error Count: ", 
            isoChrones.errors[[idx]], "]\t\t\t\t\t", sep = "")
        
        # Attempt to calculate the isochrone
        isoChrones[[idx]] <- tryCatch({
          osrmIsochrone(loc = locationsCoords[idx, ], osrm.profile = "car", breaks = breaks, res = res)
        }, error = function(e) {
          isoChrones.errors[[idx]] <- isoChrones.errors[[idx]] + 1
          cat("\n\tERROR:", conditionMessage(e), "\n\n", "[Facility Error Count: ", isoChrones.errors[[idx]], "]")
          NULL
        })
        
        # Check if isochrone was successfully calculated
        if (!is.null(isoChrones[[idx]])) {
          if (length(isoChrones[[idx]]$geometry) != n_breaks) {
            cat(paste("[Wrong number of layers: ", length(isoChrones[[idx]]$geometry), "]"))
            isoChrones[[idx]] <- NULL
          } else {
            isoChrones[[idx]]$drive_times <- factor(paste(
              isoChrones[[idx]]$isomin, "to", isoChrones[[idx]]$isomax, "mins"
            ))
          }
        }
        
        # Introduce a random sleep interval
        Sys.sleep(sample(seq(35, 65), 1))
      }
      
      cat("\r ", rep(" ", 40))
    }
    
    # Increase the resolution for the next iteration
    res <- res + 25
    
    # Break the loop if error limit is reached
    if (any(sapply(isoChrones.errors, `>`, 5))) break
  }
  
  cat("\n\n[ COMPLETE ]\n\n")
  return(isoChrones)
}

library(sf)
library(osrm)
gyn_onc_short <- gyn_onc %>% head(3)
breaks <- c(30, 60, 90, 120, 240)
gyn_onc_short <- sf::st_as_sf(gyn_onc_short, coords = c("long", "lat"), crs = 4326)
test <- isoChrone.build(gyn_onc_short, res = 10, breaks) 

plot(test["isomax"], breaks = sort(unique(c(iso$isomin, iso$isomax))))
```


```{r}
isochrones_df
```


```{r}
write.csv(isochrones_df, file = "isochrones_df.csv")
```


