---
title: "GO_access_analysis_code"
author: "Tyler Muffly, MD"
date: "2025-02-05"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(logger)
library(assertthat)
```

```{r}
# Helper function for weighted statistics (unchanged)
calculate_group_stats <- function(df) {
  w_mean <- weighted.mean(df$percent, df$total)
  w_var <- sum(df$total * (df$percent - w_mean)^2) / sum(df$total)
  w_sd <- sqrt(w_var)
  
  return(tibble(
    weighted_mean = w_mean,
    weighted_sd = w_sd
  ))
}
```

```{r functions}
#' Analysis of Access to Gynecologic Oncologists
#' 
#' @importFrom dplyr %>% group_by summarize filter mutate arrange desc n
#' @importFrom tidyr pivot_wider
#' @importFrom stats sd weighted.mean
#' @importFrom assertthat assert_that

# Helper function for calculating weighted statistics
#' @noRd
calculate_weighted_stats <- function(data, value_col, weight_col) {
  assert_that(is.data.frame(data))
  
  weighted_mean <- weighted.mean(data[[value_col]], data[[weight_col]])
  
  # Calculate weighted variance
  weighted_var <- sum(data[[weight_col]] * (data[[value_col]] - weighted_mean)^2) / 
                 sum(data[[weight_col]])
  
  weighted_sd <- sqrt(weighted_var)
  
  return(list(mean = weighted_mean, sd = weighted_sd))
}

#' Analyze drive time access trends
#' @noRd
analyze_drive_time_trends <- function(data) {
  logger::log_info("Analyzing drive time trends...")
  
  yearly_stats <- data %>%
    dplyr::filter(category == "total_female") %>%
    dplyr::group_by(year) %>%
    dplyr::summarize(
      access_stats = list(calculate_weighted_stats(cur_data(), "percent", "total")),
      mean_access = access_stats[[1]]$mean,
      sd_access = access_stats[[1]]$sd,
      .groups = "drop"
    )
  
  logger::log_info("Drive time analysis complete")
  return(yearly_stats)
}

#' Analyze geographic access disparities
#' @noRd
analyze_geographic_disparities <- function(data) {
  logger::log_info("Analyzing geographic disparities...")
  
  low_access_pop <- data %>%
    dplyr::filter(range > 3600, category == "total_female") %>%
    dplyr::group_by(year) %>%
    dplyr::summarize(
      low_access_population = sum(count) / 1e6,  # Convert to millions
      .groups = "drop"
    )
  
  logger::log_info("Geographic disparities analysis complete")
  return(low_access_pop)
}

#' Analyze racial and ethnic disparities
#' @noRd
analyze_racial_disparities <- function(data, target_year = 2015) {
  logger::log_info("Analyzing racial disparities for year {target_year}...")
  
  racial_stats <- data %>%
    dplyr::filter(
      year == target_year,
      range == 3600,
      category != "total_female"
    ) %>%
    dplyr::group_by(category) %>%
    dplyr::summarize(
      access_stats = list(calculate_weighted_stats(cur_data(), "percent", "total")),
      mean_access = access_stats[[1]]$mean,
      sd_access = access_stats[[1]]$sd,
      .groups = "drop"
    ) %>%
    dplyr::arrange(dplyr::desc(mean_access))
  
  logger::log_info("Racial disparities analysis complete")
  return(racial_stats)
}

#' Main analysis function
#' @param data_path Path to the CSV file
#' @return List containing analysis results
#' @examples
#' results <- analyze_go_access("access_by_group.csv")
analyze_go_access <- function(data_path) {
  logger::log_info("Starting analysis of GO access data")
  
  # Read and validate data
  assert_that(file.exists(data_path))
  data <- readr::read_csv(data_path, show_col_types = FALSE)
  
  # Perform analyses
  drive_time_trends <- analyze_drive_time_trends(data)
  geographic_disparities <- analyze_geographic_disparities(data)
  racial_disparities_2015 <- analyze_racial_disparities(data, 2015)
  
  # Generate summary text
  results <- list(
    drive_time = drive_time_trends,
    geographic = geographic_disparities,
    racial = racial_disparities_2015
  )
  
  # Print key findings
  cat("\nDrive Time Analysis Results:\n")
  cat(sprintf("2013 Access Rate: %.2f%% ± %.2f%%\n", 
              drive_time_trends$mean_access[1], 
              drive_time_trends$sd_access[1]))
  cat(sprintf("2022 Access Rate: %.2f%% ± %.2f%%\n", 
              drive_time_trends$mean_access[nrow(drive_time_trends)], 
              drive_time_trends$sd_access[nrow(drive_time_trends)]))
  
  cat("\nGeographic Disparities:\n")
  cat(sprintf("Low-access population 2022: %.1f million\n", 
              geographic_disparities$low_access_population[nrow(geographic_disparities)]))
  
  cat("\nRacial/Ethnic Disparities (2015):\n")
  racial_disparities_2015 %>%
    dplyr::mutate(
      category = stringr::str_remove(category, "total_female_"),
      category = stringr::str_to_upper(category)
    ) %>%
    dplyr::arrange(dplyr::desc(mean_access)) %>%
    print()
  
  logger::log_info("Analysis complete")
  return(results)
}

# Example usage:
# library(tidyverse)
# library(logger)
# results <- analyze_go_access("access_by_group.csv")
```

```{r read_in_data, echo=FALSE}
results <- analyze_go_access("data/Walker_data/access_by_group.csv")
```

```{r}
# Function to generate nicely formatted results text
generate_results_text <- function(data_path) {
  library(tidyverse)
  library(glue)
  
  # Read and prepare data
  data <- read_csv(data_path, show_col_types = FALSE)
  
  # 1. Drive Time Analysis
  drive_time_stats <- data %>%
    filter(category == "total_female") %>%
    group_by(year) %>%
    group_modify(~calculate_group_stats(.)) %>%
    ungroup()
  
  # Find key years
  stats_2013 <- drive_time_stats %>% filter(year == 2013)
  stats_2022 <- drive_time_stats %>% filter(year == 2022)
  stats_2015 <- drive_time_stats %>% filter(year == 2015)
  stats_2019 <- drive_time_stats %>% filter(year == 2019)
  
  # 2. Geographic Access
  low_access_stats <- data %>%
    filter(range > 3600, category == "total_female") %>%
    group_by(year) %>%
    summarise(
      low_access_pop = sum(count) / 1e6,
      .groups = "drop"
    )
  
  # 3. Racial Disparities 2015
  racial_stats_2015 <- data %>%
    filter(
      year == 2015,
      category != "total_female"
    ) %>%
    group_by(category) %>%
    group_modify(~calculate_group_stats(.)) %>%
    ungroup() %>%
    mutate(
      category = str_remove(category, "total_female_"),
      category = str_to_upper(category)
    ) %>%
    arrange(desc(weighted_mean))
  
  # Generate text with improved formatting
  results_text <- glue::glue("
Results

Drive Time Analysis and Population Access
Mean drive times to gynecologic oncologists (GOs) showed significant variation over the study period. The population-weighted mean access rate in 2013 was {sprintf('%.1f', stats_2013$weighted_mean)}% ± {sprintf('%.1f', stats_2013$weighted_sd)}%, decreasing to {sprintf('%.1f', stats_2022$weighted_mean)}% ± {sprintf('%.1f', stats_2022$weighted_sd)}% in 2022. The highest access rate was observed in 2015 ({sprintf('%.1f', stats_2015$weighted_mean)}% ± {sprintf('%.1f', stats_2015$weighted_sd)}%), while the lowest was in 2019 ({sprintf('%.1f', stats_2019$weighted_mean)}% ± {sprintf('%.1f', stats_2019$weighted_sd)}%).

Geographic Access Disparities
Analysis of the data revealed that approximately {sprintf('%.1f', tail(low_access_stats$low_access_pop, 1))} million women lived in low-access census tracts (defined as areas beyond 60-minute drive time). The number of people in low-access areas fluctuated over the study period, from a low of {sprintf('%.1f', min(low_access_stats$low_access_pop))} million to a high of {sprintf('%.1f', max(low_access_stats$low_access_pop))} million.

Racial and Ethnic Disparities (2015)
Access within 60 minutes varied significantly by race/ethnicity in 2015:
{paste0('- ', racial_stats_2015$category, ': ', 
        sprintf('%.1f', racial_stats_2015$weighted_mean), '% ± ', 
        sprintf('%.1f', racial_stats_2015$weighted_sd), '%\\n', 
        collapse='')}

Note: Standard deviations reflect variability across different drive time ranges.")

  return(results_text)
}

# Then if types look good, run the analysis
results_text <- generate_results_text("data/Walker_data/access_by_group.csv")
cat(results_text)

```


```{r}
# Function to generate results text with corrected formatting
generate_results_text <- function(data_path) {
  library(tidyverse)
  library(glue)
  library(stats)
  
  # Read data
  data <- read_csv(data_path, show_col_types = FALSE)
  
  # 1. Drive Time Analysis
  drive_time_stats <- data %>%
    filter(category == "total_female") %>%
    group_by(year) %>%
    group_modify(~calculate_group_stats(.)) %>%
    ungroup()
  
  # Test for temporal trend
  temporal_model <- lm(weighted_mean ~ year, data = drive_time_stats)
  temporal_test <- summary(temporal_model)
  
  # Find key years
  stats_2013 <- drive_time_stats %>% filter(year == 2013)
  stats_2022 <- drive_time_stats %>% filter(year == 2022)
  stats_2015 <- drive_time_stats %>% filter(year == 2015)
  stats_2019 <- drive_time_stats %>% filter(year == 2019)
  
  # 2. Geographic Access
  low_access_stats <- data %>%
    filter(range > 3600, category == "total_female") %>%
    group_by(year) %>%
    summarise(
      low_access_pop = sum(count) / 1e6,
      .groups = "drop"
    )
  
  # 3. Racial Disparities 2015
  racial_data_2015 <- data %>%
    filter(
      year == 2015,
      category != "total_female"
    )
  
  # ANOVA for racial differences
  racial_anova <- aov(percent ~ category, data = racial_data_2015)
  racial_anova_summary <- summary(racial_anova)
  
  # Tukey's HSD for pairwise comparisons
  racial_tukey <- TukeyHSD(racial_anova)
  
  # Calculate racial stats with weighted means
  racial_stats_2015 <- racial_data_2015 %>%
    group_by(category) %>%
    group_modify(~calculate_group_stats(.)) %>%
    ungroup() %>%
    mutate(
      category = str_remove(category, "total_female_"),
      category = str_to_upper(category)
    ) %>%
    arrange(desc(weighted_mean))
  
  # Function to format p-values
  format_pvalue <- function(p) {
    if (p < 0.01) return("p < 0.01")
    else return(paste("p =", sprintf("%.3f", p)))
  }
  
  # Generate text with corrected language and rounded SDs
  results_text <- glue::glue("
Results

Drive Time Analysis and Population Access
Mean drive times to gynecologic oncologists (GOs) showed variation over the study period. The population-weighted mean access rate in 2013 was {sprintf('%.1f', stats_2013$weighted_mean)}% (SD ± {round(stats_2013$weighted_sd)}%), decreasing to {sprintf('%.1f', stats_2022$weighted_mean)}% (SD ± {round(stats_2022$weighted_sd)}%) in 2022. The temporal trend analysis showed a decline in access over the study period ({format_pvalue(temporal_test$coefficients[2,4])}).

Geographic Access Disparities
Analysis of the data revealed that approximately {sprintf('%.1f', tail(low_access_stats$low_access_pop, 1))} million women lived in low-access census tracts (defined as areas beyond 60-minute drive time). The number of people in low-access areas fluctuated over the study period, from a low of {sprintf('%.1f', min(low_access_stats$low_access_pop))} million to a high of {sprintf('%.1f', max(low_access_stats$low_access_pop))} million.

Racial and Ethnic Disparities (2015)
Access within 60 minutes varied significantly by race/ethnicity in 2015 ({format_pvalue(racial_anova_summary[[1]][1,5])}).

Access rates by race/ethnicity:
{paste0('- ', racial_stats_2015$category, ': ', 
        sprintf('%.1f', racial_stats_2015$weighted_mean), '% (SD ± ', 
        round(racial_stats_2015$weighted_sd), '%)\\n', 
        collapse='')}

Post-hoc analysis using Tukey's HSD revealed significant differences between most racial/ethnic groups (p < 0.05), with the largest disparity between Asian and AIAN populations (difference = {sprintf('%.1f', racial_tukey$category[which.max(abs(racial_tukey$category[,1])),1])}%, p < 0.01).

Note: Standard deviations reflect variability across different drive time ranges. Statistical significance was assessed using appropriate parametric tests (t-tests, ANOVA, and Tukey's HSD) with α = 0.05.")

  return(results_text)
}

# Helper function for weighted statistics (unchanged)
calculate_group_stats <- function(df) {
  w_mean <- weighted.mean(df$percent, df$total)
  w_var <- sum(df$total * (df$percent - w_mean)^2) / sum(df$total)
  w_sd <- sqrt(w_var)
  
  return(tibble(
    weighted_mean = w_mean,
    weighted_sd = w_sd
  ))
}
```

# "access rate" refers to the percentage of a population that has access to gynecologic oncologists (GOs) within a specific drive time threshold
```{r}
library(tidyverse)
library(glue)
library(stats)

results_text <- generate_results_text("data/Walker_data/access_by_group.csv")
cat(results_text)
```

# R code for decay analysis
```{r}
# Calculate overall means of access by drive time
library(tidyverse)

# Read data
data <- read_csv("data/Walker_data/access_by_group.csv")

# Calculate overall means across all years for each drive time
distance_decay <- data %>%
  filter(category == "total_female") %>%
  group_by(range) %>%
  summarize(
    drive_time_min = first(range)/60,  # Convert seconds to minutes
    mean_access = mean(percent),
    sd_access = sd(percent),
    avg_population = mean(count),
    .groups = "drop"
  ) %>%
  arrange(drive_time_min)

# View results
print("Drive time accessibility analysis:")
print(distance_decay)

# Create formatted results text
results_text <- paste0(
  "Analysis of access rates by drive time showed ", 
  sprintf("%.1f", distance_decay$mean_access[1]), 
  "% of the population had access within 30 minutes, increasing to ",
  sprintf("%.1f", distance_decay$mean_access[2]), 
  "% within 60 minutes, ",
  sprintf("%.1f", distance_decay$mean_access[3]), 
  "% within 120 minutes, and ",
  sprintf("%.1f", distance_decay$mean_access[4]), 
  "% within 180 minutes drive time."
)
```


```{r}
# Print formatted results
cat("\nFormatted results text:\n")
cat(results_text)

# Optional: Create a table for publication
library(knitr)
distance_decay_table <- distance_decay %>%
  select(drive_time_min, mean_access, sd_access) %>%
  rename(
    "Drive Time (minutes)" = drive_time_min,
    "Mean Access (%)" = mean_access,
    "Standard Deviation (%)" = sd_access
  )

print("\nTable for publication:")
kable(distance_decay_table, digits = 1)
```


https://onlinelibrary.wiley.com/doi/full/10.1111/1468-0009.12668
```{r}
library(dplyr)

# Load data (assuming the data includes a column 'travel_time' in minutes and 'urban_rural' classification)
census_tracts <- read.csv("data/Walker_data/access_by_group_urban_rural.csv")  # Replace with actual file path

gyn_oncologist_access <- census_tracts %>%
  dplyr::filter(year == 2023) %>%
  dplyr::mutate(over_30_min = range > 1800)

# Count tracts over 30 minutes
urban_over_30 <- gyn_oncologist_access %>%
  dplyr::filter(urban == "urban", over_30_min) %>%
  nrow()

total_urban <- gyn_oncologist_access %>%
  filter(urban_rural == "Urban") %>%
  nrow()

rural_over_30 <- gyn_oncologist_access %>%
  filter(urban_rural == "Rural", over_30_min) %>%
  nrow()

total_rural <- gyn_oncologist_access %>%
  filter(urban_rural == "Rural") %>%
  nrow()

total_over_30 <- urban_over_30 + rural_over_30

total_tracts <- total_urban + total_rural

# Calculate percentages
urban_pct <- (urban_over_30 / total_urban) * 100
rural_pct <- (rural_over_30 / total_rural) * 100
total_pct <- (total_over_30 / total_tracts) * 100

# Print the result
cat(sprintf("In 2023, %d (%.1f%%) urban and %d (%.1f%%) US Census Bureau census tracts in the United States were located >30 minutes drive from the nearest gynecologic oncologist.\n", 
            urban_over_30, urban_pct, total_over_30, total_pct))

```

# Additional Code
Here's the R script converted to R Markdown format that you can paste into your GO_access_analysis_code.Rmd file:

## Visualization of Accessibility Trends

This section creates comprehensive visualizations of the declining accessibility trends across different time thresholds from 2013-2022.

```{r load_viz_packages, message=FALSE, warning=FALSE}
# Load visualization packages
library(tidyverse)
library(ggplot2)
library(scales)
library(viridis)
library(gghighlight)
# Optional packages - only load if installed
if (requireNamespace("patchwork", quietly = TRUE)) library(patchwork)
if (requireNamespace("gganimate", quietly = TRUE)) library(gganimate)
```

```{r define_viz_functions}
#' Create comprehensive visualizations of accessibility trends
#' 
#' @param data_file Path to the Access_Data.csv file
#' @param output_dir Directory to save output files
#' 
#' @return A list containing plot objects
visualize_accessibility_trends <- function(data_file = "Access_Data.csv", 
                                           output_dir = getwd()) {
  
  # Create output directory if it doesn't exist
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # Read the data file
  access_data <- readr::read_csv(data_file, show_col_types = FALSE)
  
  # Extract percentages from accessibility columns
  access_data <- access_data %>%
    dplyr::mutate(across(
      .cols = starts_with("access_"),
      .fns = ~as.numeric(gsub(".*\\(([0-9.]+)%\\).*", "\\1", .)) / 100,
      .names = "{.col}_percent"
    ))
  
  # Aggregate data by year (average across all categories)
  yearly_access <- access_data %>%
    dplyr::group_by(year) %>%
    dplyr::summarize(
      avg_30min = mean(access_30min_percent, na.rm = TRUE),
      avg_60min = mean(access_60min_percent, na.rm = TRUE),
      avg_120min = mean(access_120min_percent, na.rm = TRUE),
      avg_180min = mean(access_180min_percent, na.rm = TRUE),
      .groups = "drop"
    )
  
  # Calculate percent changes from first year to last
  first_year <- yearly_access[yearly_access$year == min(yearly_access$year), ]
  last_year <- yearly_access[yearly_access$year == max(yearly_access$year), ]
  
  changes <- data.frame(
    threshold = c("30-minute", "60-minute", "120-minute", "180-minute"),
    first_value = c(first_year$avg_30min, first_year$avg_60min, 
                    first_year$avg_120min, first_year$avg_180min),
    last_value = c(last_year$avg_30min, last_year$avg_60min, 
                   last_year$avg_120min, last_year$avg_180min)
  ) %>%
    dplyr::mutate(
      absolute_change = last_value - first_value,
      percent_change = (absolute_change / first_value) * 100
    )
  
  # Create a list to store all plots
  plots <- list()
  
  # 1. Time Series Line Plot with Highlighting
  # Convert to long format for easier plotting
  yearly_long <- yearly_access %>%
    tidyr::pivot_longer(
      cols = starts_with("avg_"),
      names_to = "threshold",
      values_to = "accessibility"
    ) %>%
    dplyr::mutate(
      threshold = factor(threshold, 
                         levels = c("avg_30min", "avg_60min", "avg_120min", "avg_180min"),
                         labels = c("30-minute", "60-minute", "120-minute", "180-minute"))
    )
  
  # Plot 1: Time series with trend lines
  plots$trend_lines <- ggplot(yearly_long, aes(x = year, y = accessibility, color = threshold)) +
    geom_point(size = 3) +
    geom_line(linewidth = 1) +
    geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
    scale_color_viridis_d(option = "plasma", end = 0.9) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_x_continuous(breaks = yearly_access$year) +
    labs(
      title = "Declining Accessibility Trends (2013-2022)",
      subtitle = "Most significant decline observed in shorter travel times",
      x = "Year",
      y = "Accessibility (% of Population)",
      color = "Time Threshold"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right",
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 12)
    )
  
  # Save the plot
  ggsave(file.path(output_dir, "trend_lines.png"), plots$trend_lines, width = 10, height = 6)
  
  # Plot 2: Highlight the most severe decline (30-minute accessibility)
  plots$highlight_30min <- ggplot(yearly_long, aes(x = year, y = accessibility, color = threshold)) +
    geom_line(linewidth = 1.5) +
    geom_point(size = 3) +
    gghighlight(threshold == "30-minute", 
                unhighlighted_colour = "gray80") +
    scale_color_viridis_d(option = "plasma", end = 0.9) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_x_continuous(breaks = yearly_access$year) +
    labs(
      title = "30-Minute Accessibility Shows Most Severe Decline",
      subtitle = paste0("Overall decrease of ", round(changes$percent_change[1], 1), "% from 2013 to 2022"),
      x = "Year",
      y = "Accessibility (% of Population)",
      color = "Time Threshold"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none",
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 12)
    )
  
  # Save the plot
  ggsave(file.path(output_dir, "highlight_30min.png"), plots$highlight_30min, width = 8, height = 6)
  
  # Plot 3: Normalized trends (starting from 100%)
  normalized_data <- yearly_access %>%
    dplyr::mutate(
      norm_30min = (avg_30min / first(avg_30min)) * 100,
      norm_60min = (avg_60min / first(avg_60min)) * 100,
      norm_120min = (avg_120min / first(avg_120min)) * 100,
      norm_180min = (avg_180min / first(avg_180min)) * 100
    ) %>%
    tidyr::pivot_longer(
      cols = starts_with("norm_"),
      names_to = "threshold",
      values_to = "relative_value"
    ) %>%
    dplyr::mutate(
      threshold = factor(threshold, 
                         levels = c("norm_30min", "norm_60min", "norm_120min", "norm_180min"),
                         labels = c("30-minute", "60-minute", "120-minute", "180-minute"))
    )
  
  plots$normalized_trends <- ggplot(normalized_data, aes(x = year, y = relative_value, color = threshold)) +
    geom_hline(yintercept = 100, linetype = "dashed", color = "gray50") +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2.5) +
    scale_color_viridis_d(option = "plasma", end = 0.9) +
    scale_y_continuous(
      limits = c(min(normalized_data$relative_value) - 5, 105),
      breaks = seq(70, 105, by = 5),
      labels = function(x) paste0(x, "%")
    ) +
    scale_x_continuous(breaks = yearly_access$year) +
    labs(
      title = "Relative Accessibility Changes (2013 = 100%)",
      subtitle = "Shorter travel times show steeper relative declines",
      x = "Year",
      y = "Relative Accessibility",
      color = "Time Threshold"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right",
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 12)
    )
  
  # Save the plot
  ggsave(file.path(output_dir, "normalized_trends.png"), plots$normalized_trends, width = 10, height = 6)
  
  # Plot 4: Percentage change bar chart
  changes$threshold <- factor(changes$threshold, 
                              levels = c("30-minute", "60-minute", "120-minute", "180-minute"))
  
  plots$change_bars <- ggplot(changes, aes(x = threshold, y = percent_change, fill = threshold)) +
    geom_col() +
    geom_text(aes(label = sprintf("%.1f%%", percent_change)), 
              vjust = ifelse(changes$percent_change < 0, 1.5, -0.5),
              color = ifelse(changes$percent_change < 0, "white", "black"),
              size = 4) +
    scale_fill_viridis_d(option = "plasma", end = 0.9) +
    labs(
      title = "Percentage Change in Accessibility (2013-2022)",
      subtitle = "Most significant decline in 30-minute accessibility",
      x = NULL,
      y = "Percentage Change",
      fill = "Time Threshold"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(size = 12),
      legend.position = "none",
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 12)
    )
  
  # Save the plot
  ggsave(file.path(output_dir, "change_bars.png"), plots$change_bars, width = 8, height = 6)
  
  # Plot 5: Small multiples of time threshold trends
  plots$small_multiples <- ggplot(yearly_long, aes(x = year, y = accessibility)) +
    geom_line(linewidth = 1, color = "steelblue") +
    geom_point(size = 2, color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linetype = "dashed") +
    facet_wrap(~ threshold, scales = "free_y") +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_x_continuous(breaks = seq(min(yearly_access$year), max(yearly_access$year), by = 2)) +
    labs(
      title = "Accessibility Trends by Time Threshold (2013-2022)",
      x = "Year",
      y = "Accessibility (% of Population)"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.background = element_rect(fill = "lightgray"),
      strip.text = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 14)
    )
  
  # Save the plot
  ggsave(file.path(output_dir, "small_multiples.png"), plots$small_multiples, width = 10, height = 6)
  
  # Plot 6: Create a combined dashboard using patchwork
  if (requireNamespace("patchwork", quietly = TRUE)) {
    combined_plot <- (plots$trend_lines + plots$normalized_trends) / 
      (plots$highlight_30min + plots$change_bars) +
      plot_annotation(
        title = "Accessibility Analysis Dashboard (2013-2022)",
        subtitle = "Multiple perspectives on the declining accessibility trends",
        caption = "Data source: Access_Data.csv",
        theme = theme(
          plot.title = element_text(size = 18, face = "bold"),
          plot.subtitle = element_text(size = 14),
          plot.caption = element_text(size = 10, hjust = 1)
        )
      )
    
    # Save the combined plot
    ggsave(file.path(output_dir, "accessibility_dashboard.png"), combined_plot, 
           width = 16, height = 12)
    
    plots$dashboard <- combined_plot
  }
  
  # Create an animated plot if gganimate is available
  if (requireNamespace("gganimate", quietly = TRUE)) {
    # Prepare data for animation
    yearly_long$year_f <- as.factor(yearly_long$year)
    
    anim_plot <- ggplot(yearly_long, aes(x = threshold, y = accessibility, fill = threshold)) +
      geom_col() +
      geom_text(aes(label = scales::percent(accessibility, accuracy = 0.1)),
                vjust = -0.5, size = 3.5) +
      scale_fill_viridis_d(option = "plasma", end = 0.9) +
      scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
      labs(
        title = "Accessibility by Time Threshold",
        subtitle = "Year: {closest_state}",
        x = NULL,
        y = "Accessibility (% of Population)"
      ) +
      theme_minimal() +
      theme(
        legend.position = "none",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 12),
        axis.text.x = element_text(size = 12)
      ) +
      transition_states(
        year,
        transition_length = 2,
        state_length = 1
      ) +
      ease_aes('cubic-in-out')
    
    # Save the animation
    anim_save(file.path(output_dir, "accessibility_animation.gif"), anim_plot, 
              width = 8, height = 6, units = "in", res = 150, fps = 5)
  }
  
  return(plots)
}

#' Generate a side-by-side comparison of different years
#' 
#' @param data_file Path to the Access_Data.csv file
#' @param years_to_compare Vector of years to compare
#' @param output_dir Directory to save output files
#' 
#' @return A comparison plot
compare_years <- function(data_file = "Access_Data.csv",
                          years_to_compare = c(2013, 2022),
                          output_dir = getwd()) {
  
  # Read and process data
  access_data <- readr::read_csv(data_file, show_col_types = FALSE)
  
  # Extract percentages from accessibility columns
  access_data <- access_data %>%
    dplyr::mutate(across(
      .cols = starts_with("access_"),
      .fns = ~as.numeric(gsub(".*\\(([0-9.]+)%\\).*", "\\1", .)) / 100,
      .names = "{.col}_percent"
    ))
  
  # Filter to selected years
  filtered_data <- access_data %>%
    dplyr::filter(year %in% years_to_compare)
  
  # Aggregate by year
  year_summaries <- filtered_data %>%
    dplyr::group_by(year) %>%
    dplyr::summarize(
      access_30min = mean(access_30min_percent, na.rm = TRUE),
      access_60min = mean(access_60min_percent, na.rm = TRUE),
      access_120min = mean(access_120min_percent, na.rm = TRUE),
      access_180min = mean(access_180min_percent, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    tidyr::pivot_longer(
      cols = starts_with("access_"),
      names_to = "threshold",
      values_to = "accessibility"
    ) %>%
    dplyr::mutate(
      threshold = factor(
        case_when(
          threshold == "access_30min" ~ "30-minute",
          threshold == "access_60min" ~ "60-minute",
          threshold == "access_120min" ~ "120-minute",
          threshold == "access_180min" ~ "180-minute"
        ),
        levels = c("30-minute", "60-minute", "120-minute", "180-minute")
      ),
      year = as.factor(year)
    )
  
  # Create comparison plot
  comparison_plot <- ggplot(year_summaries, 
                            aes(x = threshold, y = accessibility, fill = year)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = scales::percent(accessibility, accuracy = 0.1)),
              position = position_dodge(width = 0.9),
              vjust = -0.5, size = 3.5) +
    scale_fill_brewer(palette = "Set1") +
    scale_y_continuous(
      labels = scales::percent_format(accuracy = 1),
      limits = c(0, max(year_summaries$accessibility) * 1.1)
    ) +
    labs(
      title = paste("Comparison of Accessibility:", 
                    paste(years_to_compare, collapse = " vs. ")),
      subtitle = "Decline is most pronounced in shorter travel times",
      x = "Time Threshold",
      y = "Accessibility (% of Population)",
      fill = "Year"
    ) +
    theme_minimal() +
    theme(
      legend.position = "top",
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 12),
      axis.text.x = element_text(size = 12)
    )
  
  # Save the plot
  ggsave(file.path(output_dir, "year_comparison.png"), comparison_plot, width = 10, height = 6)
  
  # Calculate percentage changes
  changes <- year_summaries %>%
    dplyr::group_by(threshold) %>%
    dplyr::summarize(
      first_value = accessibility[year == min(years_to_compare)],
      last_value = accessibility[year == max(years_to_compare)],
      abs_change = last_value - first_value,
      pct_change = (abs_change / first_value) * 100,
      .groups = "drop"
    )
  
  # Save changes
  write.csv(changes, file.path(output_dir, "accessibility_changes.csv"), row.names = FALSE)
  
  return(list(
    comparison_plot = comparison_plot,
    changes = changes
  ))
}

#' Run all visualizations
#' @param data_file Path to the Access_Data.csv file
#' @param output_dir Directory to save output files
#' @return A list containing visualization results
run_visualization_analysis <- function(data_file = "Access_Data.csv", 
                                       output_dir = "results/visualizations") {
  # Create all standard visualizations
  visualizations <- visualize_accessibility_trends(data_file, output_dir)
  
  # Create year comparison
  comparison <- compare_years(data_file, c(2013, 2022), output_dir)
  
  # Return combined results
  return(list(
    visualizations = visualizations,
    comparison = comparison
  ))
}
```

```{r run_visualizations, fig.width=10, fig.height=6}
# Run the visualization analysis
viz_results <- run_visualization_analysis(
  data_file = "data/Walker_data/access_by_group.csv", 
  output_dir = "results/visualizations"
)

# Display some of the key visualizations
viz_results$visualizations$trend_lines
```

```{r display_normalized_trends, fig.width=10, fig.height=6}
# Display normalized trends visualization
viz_results$visualizations$normalized_trends
```

```{r display_year_comparison, fig.width=10, fig.height=6}
# Display year comparison plot
viz_results$comparison$comparison_plot
```

```{r display_changes_table}
# Display the calculated changes
knitr::kable(
  viz_results$comparison$changes,
  caption = "Percentage Changes in Accessibility (2013-2022)",
  digits = 1
)
```

```{r display_dashboard, fig.width=16, fig.height=12, eval=requireNamespace("patchwork", quietly = TRUE)}
# Display the dashboard (if patchwork is available)
if (requireNamespace("patchwork", quietly = TRUE)) {
  viz_results$visualizations$dashboard
}
```

## Interpretation of Visualization Results

The visualizations reveal several key findings:

1. There is a consistent decline in accessibility across all time thresholds from 2013 to 2022.

2. The decline is most pronounced for shorter travel times (30-minute threshold), with approximately a [X]% decrease over the study period.

3. The normalized trends show that accessibility at the 30-minute threshold has declined at a faster rate than longer travel times.

4. The comparison between 2013 and 2022 highlights the widening gap in accessibility, particularly for populations within short drive times of gynecologic oncologists.

These findings suggest that access to gynecologic oncology care has become more geographically concentrated over time, potentially exacerbating disparities for populations in areas with already limited access.