---
title: "isochrones"
author: "Tyler Muffly, MD"
date: "2024-08-31"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source("/Users/tylermuffly/Dropbox (Personal)/isochrones/R/01-setup.R")
library(dplyr)
library(tidycensus)
library(sf)
library(purrr)
library(lubridate)
library(tidycensus)
library(dplyr)
library(lme4)
library(DT)

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(lubridate::year)
conflicted::conflicts_prefer(exploratory::str_remove_all)
conflicted::conflicts_prefer(dplyr::recode)

options(tigris_use_cache = TRUE)
```


```{r}
knitr::opts_chunk$set(
  message=FALSE, echo=FALSE, include = TRUE, fig.width = 6, fig.height = 4, progress = TRUE, warning = FALSE, collapse = FALSE)

options(scipen=999)
```

# Dates and Times For Isochrones
```{r, include = FALSE}
destination_dir <- "/Volumes/Video Projects Muffly 1/isochrones_data/walker_script"
iso_datetime_yearly <- tibble(
  date = c("2013-10-18 09:00:00", "2014-10-17 09:00:00", "2015-10-16 09:00:00",
           "2016-10-21 09:00:00", "2017-10-20 09:00:00", "2018-10-19 09:00:00",
           "2019-10-18 09:00:00", "2020-10-16 09:00:00", "2021-10-15 09:00:00",
           "2022-10-21 09:00:00", "2023-10-20 09:00:00"),
  year = c("2013", "2014", "2015", "2016", "2017", "2018", "2019", "2020", "2021", "2022", "2023")
)
```

# Physician Locations
```{r, include = TRUE}
# Let's build the physician location table; presumably you have a separate table you can plug in
# for this step.
# Your physician locations data
physician_locations <- data.frame(
  name = c("Alpha, MD", "Beta, MD", "Delta, MD", "Gamma, MD", "Zulu, MD", "Xray, MD", "Yankee, MD", "Alpha, MD"),
  Latitude = c(39.7392, 41.8781, 40.7128, 39.7294, 42.3314, 48.7519, 18.4655, 39.7392),
  Longitude = c(-104.9903, -87.6298, -74.0060, -104.8319, -83.0458, -122.4787, -66.1057, -104.9903),
  year = c(2023, 2023, 2023, 2023, 2023, 2023, 2023, 2023)
)
```

# Create Isochrones from Physician Locations
```{r, include = FALSE, eval = FALSE}
# We'll want to convert our table to an sf object to calculate the isochrones.
physician_sf <- physician_locations %>%
  st_as_sf(coords = c("Longitude", "Latitude"),
           crs = 4326)

# Function to calculate isochrones for a specific year
calculate_isochrones <- function(date, year, physician_sf) {
  log_message <- paste("Calculating isochrones for year:", year)
  print(log_message)
  
  result <- physician_sf %>%
    dplyr::filter(year == year) %>%
    isoline(
      datetime = as.POSIXct(date, tz = "UTC"),
      range = c(30, 60, 120, 180) * 60
    ) %>%
    dplyr::mutate(year = year)
  
  return(result)
}

# Iterate over each year and calculate isochrones
physician_isos_all_years <- iso_datetime_yearly %>%
  purrr::pmap_dfr(~calculate_isochrones(..1, ..2, physician_sf))

# Extract the year from the departure column and add it as a new column
physician_isos_all_years <- physician_isos_all_years %>%
  dplyr::mutate(Year = year(ymd_hms(departure)))

# Check the resulting combined sf object
print(physician_isos_all_years)

# physician_isos_all_years <- read_rds("data/physician_isos_all_years.rds")
write_rds(physician_isos_all_years, "data/physician_isos_all_years.rds")

#physician_isos_all_years <- read_rds("data/physician_isos_all_years.rds")
```

```{r, include = TRUE}
physician_isos_unique <- physician_isos_all_years %>%
  group_by(year, range) %>%
  summarise()
```

# Interactive Map of Physicians
```{r, include = TRUE}
# Adjust the number of colors to match the number of unique years
unique_years <- length(unique(physician_isos_unique$year))
unique_range <- length(unique(physician_isos_unique$range))
physician_isos_all_years <- physician_isos_unique %>%
  select(-Year) %>%
  arrange(year, range)

fancy_map <- mapview(
  physician_isos_all_years,
  zcol = "range",
  burst = TRUE,
  legend = TRUE,
  layer.name = "Physician ISOs",
  alpha.regions = 0.6,
  col.regions = viridis::viridis(unique_range), # Adjust number of colors
  homebutton = TRUE,
  basemaps = c("CartoDB.Positron", "OpenStreetMap", "Esri.WorldImagery")
); fancy_map
```

```{r cars}
# Function to pull ACS data for each year listed in iso_datetime_yearly and process it
pull_acs_data_by_year <- function(iso_datetime_yearly, all_vars, destination_dir, geography = "county") {
  
  # Set tigris to use cache
  options(tigris_use_cache = TRUE)
  
  # Ensure the destination directory exists
  if (!dir.exists(destination_dir)) {
    dir.create(destination_dir, recursive = TRUE)
    message(paste("Created directory:", destination_dir))
  }
  
  # Create an empty list to store the results
  results_list <- list()
  
  # Iterate over each row in iso_datetime_yearly to pull ACS data for each year
  for (i in seq_along(iso_datetime_yearly$year)) {
    current_year <- as.integer(iso_datetime_yearly$year[i])
    
    log_message <- paste("Starting to pull ACS data for year:", current_year, "and geography:", geography)
    print(log_message)
    
    # Try to pull ACS data for the current year
    acs_data <- tryCatch({
      get_acs(
        geography = geography,
        variables = all_vars,
        state = c(state.abb, "DC", "PR"), # Grab for the US
        year = current_year,
        geometry = TRUE,
        output = "wide"
      )
    }, error = function(e) {
      error_message <- paste("Error retrieving data for year:", current_year, "-", e$message)
      message(error_message)
      return(NULL)
    })
    
    # Skip if the data retrieval failed
    if (is.null(acs_data)) {
      message(paste("Skipping year", current_year, "due to errors"))
      next
    }
    
    # Add the year column to the data
    acs_data <- acs_data %>%
      mutate(year = current_year)
    
    log_message <- paste("Successfully pulled ACS data for year:", current_year, "and geography:", geography)
    print(log_message)
    
    # Append the data to the results list
    results_list[[i]] <- acs_data
  }
  
  # Combine all the results into a single data frame
  combined_data <- dplyr::bind_rows(results_list)
  
  # Define the output file path
  timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
  output_file <- file.path(destination_dir, paste0("combined_acs_data_", geography, "_", timestamp, ".rds"))
  
  # Save the combined data to a file
  tryCatch({
    saveRDS(combined_data, output_file)
    message(paste("Data successfully saved to:", output_file))
  }, error = function(e) {
    error_message <- paste("Failed to save data to file:", output_file, "-", e$message)
    message(error_message)
  })
  
  return(combined_data)
}
```

```{r pressure, echo=FALSE}
combined_acs_data <- pull_acs_data_by_year(iso_datetime_yearly, all_vars, destination_dir, geography = "county")

combined_acs_data1 <- combined_acs_data %>%
  select(!ends_with("M")) %>%
  rename_with(.fn = ~stringr::str_remove(.x, "E$")) %>%
  st_transform("ESRI:102003"); combined_acs_data1

readr::write_rds(combined_acs_data1, "/Volumes/Video Projects Muffly 1/isochrones_data/walker_script/acs_data_us_all_years.rds")

#combined_acs_data1 <- readr::read_rds("/Volumes/Video Projects Muffly 1/isochrones_data/walker_script/acs_data_us_all_years.rds")
```

All of USA
```{r}
years <- 2013:2022 # Replace with the years you're interested in

# Function to pull ACS data for a given year and geography
pull_acs_data_for_year <- function(year, variables, geography) {
  get_acs(
    geography = geography,
    variables = variables,
    year = year,
    survey = "acs5"
  ) %>%
    mutate(year = year, geography = geography)
}

# Initialize an empty dataframe
acs_data_us_all_years <- tibble()

# Loop through each year and bind the results
for (year in years) {
  acs_data_year <- pull_acs_data_for_year(year, all_vars, geography = "us")
  acs_data_us_all_years <- dplyr::bind_rows(acs_data_us_all_years, acs_data_year)
}; acs_data_us_all_years


combined_acs_data1 <- combined_acs_data %>%
  select(!ends_with("M")) %>%
  rename_with(.fn = ~stringr::str_remove(.x, "E$")) %>%
  sf::st_transform("ESRI:102003"); combined_acs_data1

write_rds(combined_acs_data1, "/Volumes/Video Projects Muffly 1/isochrones_data/walker_script/acs_data_us_all_years.rds")

```

ACS Variables That were Downloaded with tidycensus
```{r}
DT::datatable(head(combined_acs_data1, 100))
```

The st_centroid(combined_acs_data1) function is used to calculate the geometric centroid (center point) of each feature in a spatial object. In this case, combined_acs_data1 is a spatial data frame (likely an sf object), which contains geographic features represented as polygons.  Gets the ACS data ready for mapping.  
```{r}
centroids_2020 <- st_centroid(combined_acs_data1)
```

```{r}
# List to store tabulated results for all years
tabulated_all_years_list <- list()

# Loop through each year present in physician_isos_unique
for (current_year in unique(physician_isos_unique$year)) {
  
  # Filter data for the current year
  centroids_year <- combined_acs_data1 %>%
    filter(year == current_year)
  
  physician_isos_year <- physician_isos_unique %>%
    filter(year == current_year)
  
  # Transform, join, and keep the year column
  tabulated_year <- physician_isos_year %>%
    st_transform("ESRI:102003") %>%
    st_join(centroids_year, left = FALSE) %>%
    st_drop_geometry() %>%
    mutate(year = current_year) %>%  # Ensure the year column is present
    group_by(year, range) %>%
    summarize(across(total_female_026:total_female_hipi, .fns = ~sum(.x, na.rm = TRUE))) %>%
    ungroup()
  
  # Add the result to the list
  tabulated_all_years_list[[current_year]] <- tabulated_year
}

# Combine the results into a single dataframe
tabulated_all_years <- dplyr::bind_rows(tabulated_all_years_list)

readr::write_rds(tabulated_all_years, "/Volumes/Video Projects Muffly 1/isochrones_data/walker_script/county_centroid_tabulated_2020.rds")
```

